---
title: "Download Ensembl annotations"
author: "Chris Stubben"
date: '`r gsub("  ", " ", format(Sys.time(), "%B %e, %Y"))`'
output: github_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, comment="# ", collapse=TRUE)
```

The [hciRdata] package includes feature annotations from the latest Ensembl
release 98 for twelve reference genomes (human, mouse, elephant, fly, pig,
rabbit, rat, sheep, vervet, worm, yeast and zebrafish), plus earlier even
numbered releases for mouse and human (human90, human92, etc). Load the package
and select an annotation table.


```{r fly98, echo=-1}
options(width=110)
library(hciRdata)
dplyr::select(human98, 1:4,8)
```

To download additional species, use `read_biomart` to view a list of all available datasets.

```{r datasets, eval=FALSE}
x <- read_biomart(list="datasets")
Using Ensembl release 98
Downloaded 212 datasets
> x
# # A tibble: 212 x 3
#   dataset                      description                                  version
#   <chr>                        <chr>                                        <chr>
# 1 abrachyrhynchus_gene_ensembl Pink-footed goose genes (ASM259213v1)        ASM259213v1
# 2 acalliptera_gene_ensembl     Eastern happy genes (fAstCal1.2)             fAstCal1.2
# 3 acarolinensis_gene_ensembl   Anole lizard genes (AnoCar2.0)               AnoCar2.0
# 4 acitrinellus_gene_ensembl    Midas cichlid genes (Midas_v5)               Midas_v5
# 5 ahaastii_gene_ensembl        Great spotted kiwi genes (aptHaa1)           aptHaa1
# 6 amelanoleuca_gene_ensembl    Panda genes (ailMel1)                        ailMel1
# 7 amexicanus_gene_ensembl      Mexican tetra genes (Astyanax_mexicanus-2.0) Astyanax_mexicanus-2.0
# # … with 205 more rows
```

Use the first letter of the genus and species name in the first column to
download features (common names are supported for the reference genomes with
existing release 98 tables).

```{r panda, eval=FALSE}
panda98 <- read_biomart("amelanoleuca")
# Using Ensembl release 98
# Downloaded 23262 features
```



To download earlier versions, check the Ensembl archives for available releases.
Archived versions are available for 54, 67, and all versions after 77.

```{r listarchives, eval=FALSE}
biomaRt::listEnsemblArchives()
#              name     date                                url version current_release
# 1  Ensembl GRCh37 Feb 2014          http://grch37.ensembl.org  GRCh37
# 2      Ensembl 98 Sep 2019 http://sep2019.archive.ensembl.org      98               *
# 3      Ensembl 97 Jul 2019 http://jul2019.archive.ensembl.org      97
# ...
# 22     Ensembl 78 Dec 2014 http://dec2014.archive.ensembl.org      78
# 23     Ensembl 77 Oct 2014 http://oct2014.archive.ensembl.org      77
# 24     Ensembl 75 Feb 2014 http://feb2014.archive.ensembl.org      75
# 25     Ensembl 67 May 2012 http://may2012.archive.ensembl.org      67
# 26     Ensembl 54 May 2009 http://may2009.archive.ensembl.org      54
```



Add the version number to download earlier releases.

```{r zb87, eval=FALSE}
zb87 <- read_biomart("zebrafish", version=87)
# Downloaded 32266 features
```

In some cases, the default attributes may be missing. For example, the
`external_gene_name` is missing from early fly releases.  Use the list option
to view the available attributes.



```{r fly67_v1, eval=FALSE}
fly67 <- read_biomart("fly", version=67)
# Error in biomaRt::getBM(attributes = c("ensembl_gene_id", "external_gene_name",  :
#   Invalid attribute(s): external_gene_name
# Please use the function 'listAttributes' to get valid attribute names
fly67_columns <- read_biomart("fly", version=67, list="attributes")
fly67_columns
#    name                   description                page
#  1 ensembl_gene_id        Ensembl Gene ID            feature_page
#  2 ensembl_transcript_id  Ensembl Transcript ID      feature_page
#  3 ensembl_peptide_id     Ensembl Protein ID         feature_page
#  4 description            Description                feature_page
# ...
# 13 external_gene_id       Associated Gene Name       feature_page
# ...
```


Replace the default `external_gene_name` with `external_gene_id` by passing a
vector of attributes and then follow the code in `read_biomart` to update column
names and descriptions.

```{r fly96, eval=FALSE}
fly67 <- read_biomart("fly", version=67, attributes = c('ensembl_gene_id',
    'external_gene_id', 'gene_biotype', 'chromosome_name', 'start_position',
	 'end_position','strand', 'description', 'transcript_count'))
names(fly67)[1:6] <- c("id", "gene_name", "biotype", "chromosome", "start", "end")
fly67$description <- gsub(" \\[Source.*", "", fly67$description)
```

<br>

[hciRdata]: https://github.com/HuntsmanCancerInstitute/hciRdata
