---
title: "DESeq analysis"
output: html_document
params:
   counts: counts
   samples: samples
   annotations: hsa
   condtion: trt
   all_vs: all
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, comment="# ", collapse=TRUE)

library(hciR)

```

This guide follows the [Bioconductor RNA-Seq workflow] to find differentially expressed
genes using [DESeq2] and the [hciR] package on Github to simplify the [R] code.

### Load samples and counts

Load the samples with columns from the Sample Sheet on the Experimental Design tab in GenomEx .

```{r load_sampless, message=FALSE}
samples <- read_tsv(param$samples)
samples
```
<br>

Next, load the combined `featureCount` output files and spread into a matrix with 49671 rows and 10 samples.

```{r load_counts, message=FALSE}
count_tbl <- read_tsv("counts.txt", col_names = c("sample", "id", "count"))
count_tbl <-  spread(count_tbl, sample, count)
count_tbl
```

Reorder columns in the count matrix to match the sample table (required by DESeq) and remove 28186 features
with zero counts and 11825 features with fewer than 10 reads in any sample to create a final count matrix with 9660 rows.

```{r filter_counts}
count_tbl <- sort_counts(count_tbl, samples)
counts <- filter_counts(count_tbl, n=10)
```

### Run DESeq

Combine the samples and counts to create a `DESeqDataSet` object and run `DESeq`.


```{r DESeqobject, eval=FALSE}
dds <- DESeqDataSetFromMatrix(countData = counts,
                                colData = samples,
                                 design = ~Trt)
dds <- DESeq(dds)
rld1 <-  rlog(dds)
```


### PCA plot

Plot the first two principal components using the rlog values from the top 500 variable genes.

```{r pcaplot}
hc_plotPCA(rld1, "Trt", width=700)
```


Cluster all the rlog values using the R function `dist` to calculate the Euclidean distance between samples.
Like the PCA plot, sample 13917X9 is an outlier.

```{r d3dist}
d3heatmap_dist(rld1)
```

### Drop outliers

Drop sample 13917X9 and re-run `DESeq`.

```{r dropX9}
dds <- DESeqDataSetFromMatrix(countData = counts[, colnames(counts)!="13917X9"] ,
                                colData = samples[samples$ID!="13917X9",],
                                 design = ~Trt)
dds <- DESeq(dds)
rld <-  rlog(dds)
```

### DESeq results

Get the DESeq results using a 5% false discovery rate (FDR).

```{r DESeq}
res <- results_all(dds,  alpha= 0.05)
```

### Gene annotations

Load the mouse annotations from Ensembl.

```{r biomaRt}
data(mmu)
```


### Browse results

Combine the annotations and top 3000 genes sorted by the adjusted p-value
and create a [flex dashboard] in [DESeq_flex.html].  The [MA-plot] and [volcano plot] are linked,
so you can click and drag to create a box to highlight points and then view matching rows in the table.
You can also drag the box around the plot to easily highlight other points and rows.
In addition, you can search for genes in the table using the search box, but need to click on the row in order to
highlight the points in the plots.  The sliders can also be used to limit the results.


### Save results

Save the DESeq results to a single Excel file in [DESeq.xlsx].  The `write_DESeq` function will also
add gene annotations to the result table and output raw counts, rlog values and normalized counts.

```{r write, eval=FALSE}
write_DESeq( dds, res, rld, mmu)
```

### Plot heatmaps

Select the top 40 genes sorted by p-value and cluster the rlog differences,
so values in the heatmap represent the amount a gene deviates in a specific sample from the gene’s average across all samples.
This heatmap was created using [d3heatmap], so you can click and drag to select rows and zoom
into specific regions.   A static version using the [pheatmap] package is available in [top40_heatmap.pdf], although with slightly different
re-ordering of the dendrogram nodes.


<br>

```{r res1, echo=FALSE, fig.height=5}
# pheatmap_top(res[[1]], rld, mmu, top=40, border=NA, fontsize_row=8)
## OR interactive d3heatmap?
d3heatmap_top(res[[1]], rld, mmu, top=40, show_grid=FALSE, xaxis_font_size = 12)
```
<br>

This final heatmap displays the top 200 significant genes (zoom in to better view labels if needed).

```{r res3, echo=FALSE, fig.height=8}
d3heatmap_top(res[[1]], rld, mmu, top=200, show_grid=FALSE, xaxis_font_size = 12)
```

<br>


[R]: https://www.r-project.org
[Bioconductor RNA-Seq workflow]: http://www.bioconductor.org/help/workflows/rnaseqGene
[DESeq2]: http://www.bioconductor.org/packages/release/bioc/html/DESeq2.html
[featureCounts]: http://bioinf.wehi.edu.au/featureCounts/
[hciR]: https://github.com/HuntsmanCancerInstitute/hciR
[d3heatmap]: http://www.htmlwidgets.org/showcase_d3heatmap.html
[pheatmap]: https://cran.r-project.org/web/packages/pheatmap/index.html
[flex dashboard]: http://rmarkdown.rstudio.com/flexdashboard/
[MA-plot]: https://en.wikipedia.org/wiki/MA_plot
[volcano plot]: https://en.wikipedia.org/wiki/Volcano_plot_(statistics)
[DESeq_flex.html]: `r analysis_dir`/DESeq_flex.html
[DESeq.xlsx]: `r analysis_dir`/DESeq.xlsx
[top40_heatmap.pdf]: `r analysis_dir`/top40_heatmap.pdf
